---
title: "1. Introduction to data types"
author: "Andrew W. Park"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
library(tidyverse)
library(magrittr)
```

## Assigning data

Data come in a variety of types: for example numbers and words. First, we'll develop a repertoire of common data types before thinking about how best to organize our data. We can store data by assigning it to a variable name using the assignment operator `<-`. Modify the code so that the number `2` is assigned to the variable `a`. Then on the next line ask R to add the number `3` to `a` to check we assigned the data correctly.

```{r assignNum, exercise=TRUE, exercise.eval=FALSE}
2
```

```{r assignNum-solution}
a <- 2
a + 3
```

<div id="filter-hint">
**Hint:** Start with the variable name, then the operator, then the data to be assigned.
</div>

###

## Simple manipulation of numeric data

We can add, remove and modify data as well as apply built-in functions appropriate to a data type. To store more than one element of the same data type in a variable name, the simplest is to the use the `combine` function, which is called using just the letter `c`. We refer to these data as vectors. Here we're combine what the numbers `3` and `1` with what we already had stored in `a`, and storing the result in `b`

```{r extendVectorExample, exercise=TRUE, exercise.eval=TRUE}
b <- c(a,3,1)
```

Try introducing a new element, the number `4`, to the vector b. On the next line, ask R to return b.

```{r extendVector1, exercise=TRUE, exercise.eval=FALSE}

```


```{r extendVector1-solution}
b <- c(b,4)
b
```

We can apply built-in functions to a vector of data, for example the `sum`.

```{r sumExample, exercise=TRUE, exercise.eval=FALSE}
sum(b)
```


###


## Simple manipulation of text data

Not all data is numeric. The most common other data type is text. In R this is referred to as `character` data. 

```{r}
days <- c("Monday", "Tuesday", "Wednesday")
```







